Conceptualizing open health data platforms for low- and middle income countries

### 2024-08-13 index.tex
wip

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Antragama Abbas; Mark de Reuver},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[8]{Antragama Abbas%
%
}

\author[8]{Mark de Reuver%
%
}

\affiliation[1]{organization={PharmAccess
Foundation},city={Amsterdam},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Eindhoven University of
Technology},city={Eindhoven},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[3]{organization={PharmAccess
Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[4]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[5]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[6]{organization={Jembi Health
Systems},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}
\affiliation[7]{organization={Australian e-Health Research
Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
\affiliation[8]{organization={Delft University of
Technology},city={Delft},country={the
Netherlands},countrysep={,},postcodesep={}}

@@

It is a widely held belief that digital technologies have an important
role to play in strengthening health systems in low- and middle income
countries (LMICs), as exemplified by the WHO global strategy on digital
health \citep{who2021global}. The adoption rate of mobile phones in
LMICs has been an important driver in implementing digital health
solutions \citep{mccool2022mobile}. Yet, there are many shortcomings and
challenges, including the current fragmentation of digital platforms and
the lack of clear-cut pathways of scaling up digital health programmes,
such that they can support sustainable and equitable change of national
health systems in LMICs
\citep{mccool2022mobile, who2019recommendations, neumark2021digital}.

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
into with the proposed research agenda by de Reuver et al.~to consider
data platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. From an economic perspective, for
example, both types exhibit multi-sided markets. At the same time, data
platforms differ as their main offerings revolve around data. From an
ecosystem perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is
the conceptualization of openness. The shift in perspective from digital
platforms to data platforms coincides with the paradox of open
\citep{keller2021paradox}. Originally, openness of digital platforms
focused on open source, open standards and copyrights, which by has been
superseded by ``\ldots{} conflicts about privacy, economic value
extraction, the emergence of artificial intelligence, and the
destabilizing effects of dominant platforms on (democratic) societies.
Instead of access to information, the control of personal data has
emerged in the age of platforms as the critical contention.''
\citep{keller2021paradox}. These conflicts are particularly salient in
the healthcare domain, where people are generally willing to share their
health data to receive the best care (primary use), while the attitude
towards secondary use of health data varies greatly depending on the
type and context \citep{cascini2024health}. The shift in perspective
from digital platforms supporting primary data sharing toward data
platforms supporting secondary data sharing is one of the key issues
surrounding the polemic of data spaces \citep{otto2022designing} and
data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
Openness is particularly relevant if we are to realize a
solidarity-based approach to health data sharing that i) gives people a
greater control over their data as active decision makers; ii) ensures
that the value of data is harnessed for public good; and iii) moves
society towards equity and justice by counteracting dynamics of data
extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of open, solidarity-based HDPs
in LMICs. First, the OpenHIE framework \citep{openhie} has been adopted
by many sub-Saharan African countries \citep{mamuye2022health} as the
architectural blueprint for implementing nation-wide health information
exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
countries have, as a matter of course, extended the framework to include
``data \& analytics services'' as an additional domain. The rationale
for this addition is to facilitate secondary reuse of health data for
academic research, real-world evidence studies etc. which can be framed
within the context of ongoing efforts towards Findable, Accessible,
Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms for primary data sharing
(the original OpenHIE specification) to health data platforms for
secondary data sharing. This is problematic because the notion of
openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for secondary
data sharing and reuse. The lack of detailed specifications and
consensus of this addition to OpenHIE currently stands in the way of
development projects that aim to establish HDPs in LMICs. The purpose of
this paper, therefore, is to specificy how new standards and
technologies can be integrated into the OpenHIE architecture such that
an open HDP can be realized that supports the following different types
of data sharing.

@@

In this paper, we present a design that extends the OpenHIE
specification to include the four types of data sharing mentioned above.
Using the full-STAC approach \citep{mehl2023fullstac} we combine open
standards, open technologies and open architectures into a coherent
modular HDP platform that can be configured and reused across a variety
of use-cases. We employ a formative, naturalistic evaluation to assess
the technical risk and efficacy of the design \citep{venable2016feds}.
Given that it is prohibitively expensive to evaluate the design a
real-world setting, we aim to minimize technological risks and maximize
the efficacy of the design by considering three examples of health data
platforms in LMICs, namely:

@@

\begin{itemize}
\tightlist
\item
  What is the level of openness of the implementation, specifically in
  terms of the four types of data sharing and platform-to-platform
  opennenss?
\item
  How are the core maintenance functions of the HDP implemented, where
  we particularly focus on functions related to metadata management,
  data lineage and access control mechanisms?
\item
  Is the solution suitable for downward scalability, where we focus on
  the computer and storage requirements of the solution and assess the
  minimal requirements for on-site resources (computers, edge devices)
\end{itemize}

@@

@@

\subsection{OpenHIM platform}\label{openhim-platform}

@@

\subsection{ONA Canopy}\label{ona-canopy}

TO DO: add section describing ONA Canopy.

\subsubsection{Level of opennness}\label{level-of-opennness-1}

\subsubsection{Core maintenance
functions}\label{core-maintenance-functions-1}

\subsubsection{Downward scaleability}\label{downward-scaleability-1}

\subsection{Momcare programme}\label{momcare-programme}

@@

Clearly, a more open and scaleable platform was required if MomCare was
to be implemented in more regions. This need led to a redesign of the
underlying technical infrastructure of the MomCare project. The
objectives of this work were in fact to demonstrate a solution design
that could support the first three types of data sharing listed in
\textbf{?@tbl-types-data-sharing}. First, to investigate the viability
of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
to assess the complexity and effort required to implement the facade
pattern to integrate the legacy system into the FHIR data standard.
Using the longitudinal dataset from approximately 28 thousand patient
records, FHIR transformations script were developed and deployed using
the mediator function of the IOL. The data was transformed into 10 FHIR
v4 resources and the conceptual data model of the existing MomCare app
could readily be transformed into the FHIR standard using SQL and
validated with a Python library \citep{islam2023fhir}. The largest
challenge during the transformation process pertained to the absence of
unique business identifiers for patients and healthcare organizations.
For patients, either the mobile phone number or the healthcare insurance
number was taken, depending on availability. A combination of name,
address and latitude/longitude coordinates were used to uniquely
identify organizations and locations, as Tanzania does not have a system
in place for this purpose.

@@

\subsubsection{Level of opennness}\label{level-of-opennness-2}

@@

\subsubsection{Core maintenance
functions}\label{core-maintenance-functions-2}

@@

\subsubsection{Downward scaleability}\label{downward-scaleability-2}

@@

We specifically address the notion of openness of HDPs in LMICs in terms
of the design-related questions put forward by de Reuver at al.11:

\begin{itemize}
\tightlist
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different HDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\item
  Answers/insights to above:

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ACID & Atomicity, Consistency, Isolation, and Durability \\
CLI & Command-line Interface \\
CR & Client Registry \\
ELK & Elasticsearch, Logstach and Kibana stack \\
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
HDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
MPC & Multiparty Computation \\
PET & Privacy-enhancing technologies \\
SHR & Shared Health Record \\
\end{longtable}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\affiliation[1]{organization={PharmAccess
Foundation},city={Amsterdam},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Eindhoven University of
Technology},city={Eindhoven},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[3]{organization={PharmAccess
Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[4]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[5]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[6]{organization={Jembi Health
Systems},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}
\affiliation[7]{organization={Australian e-Health Research
Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}

@@

It is a widely held belief that digital technologies have an important
role to play in strengthening health systems in low- and middle income
countries (LMICs), as exemplified by the WHO global strategy on digital
health \citep{who2021global}. The adoption rate of mobile phones in
LMICs has been an important driver in implementing digital health
solutions \citep{mccool2022mobile}. Yet, there are many shortcomings and
challenges, including the current fragmentation of digital platforms and
the lack of clear-cut pathways of scaling up digital health programmes,
such that they can support sustainable and equitable change of national
health systems in LMICs
\citep{mehl2023fullstac, mccool2022mobile, who2019recommendations, neumark2021digital}.

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
into with the proposed research agenda by de Reuver et al.~to consider
data platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. From an economic perspective, for
example, both types exhibit multi-sided markets. At the same time, data
platforms differ as their main offerings revolve around data. From an
ecosystem perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity
\citep{dereuver2022openness}.

Particularly relevant in the context of open health data platforms
(OHDPs), is the conceptualization of openness. The shift in perspective
from digital platforms to data platforms coincides with the paradox of
open \citep{keller2021paradox}. Originally, openness of digital
platforms focused on open source, open standards and copyrights, which
by has been superseded by ``\ldots{} conflicts about privacy, economic
value extraction, the emergence of artificial intelligence, and the
destabilizing effects of dominant platforms on (democratic) societies.
Instead of access to information, the control of personal data has
emerged in the age of platforms as the critical contention.''
\citep{keller2021paradox}. These conflicts are particularly salient in
the healthcare domain, where people are generally willing to share their
health data to receive the best care (primary use), while the attitude
towards secondary use of health data varies greatly depending on the
type and context \citep{cascini2024health}. The shift in perspective
from digital platforms supporting primary data sharing toward data
platforms supporting secondary data sharing is one of the key issues
surrounding the polemic of data spaces \citep{otto2022designing} and
data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond, purtova2023data}.
Openness is particularly relevant if we are to realize a
solidarity-based approach to health data sharing that i) gives people a
greater control over their data as active decision makers; ii) ensures
that the value of data is harnessed for public good; and iii) moves
society towards equity and justice by counteracting dynamics of data
extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of open, solidarity-based
OHDPs in LMICs. First, the OpenHIE framework \citep{openhie} has been
adopted by many sub-Saharan African countries \citep{mamuye2022health}
as the architectural blueprint for implementing nation-wide health
information exchanges (HIE), including Nigeria \citep{dalhatu2023paper},
Kenya \citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}.
These countries have, as a matter of course, extended the framework to
include ``data \& analytics services'' as an additional domain. The
rationale for this addition is to facilitate secondary reuse of health
data for academic research, real-world evidence studies etc. which can
be framed within the context of ongoing efforts towards Findable,
Accessible, Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms for primary data sharing
(the original OpenHIE specification) to health data platforms for
secondary data sharing. This is problematic because the notion of
openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for secondary
data sharing and reuse. The lack of detailed specifications and
consensus of this addition to OpenHIE currently stands in the way of
development projects that aim to establish OHDPs in LMICs. The purpose
of this paper, therefore, is to specificy how new standards and
technologies can be integrated into the OpenHIE architecture such that
an open OHDP can be realized that supports the following different types
of data sharing.

@@

In this paper, we present a design that extends the OpenHIE
specification to include the three types of data sharing mentioned
above. Using the full-STAC approach \citep{mehl2023fullstac} we combine
open standards, open technologies and open architectures into a coherent
modular OHDP that can be configured and reused across a variety of
use-cases. We employ a formative, naturalistic evaluation to assess the
technical risk and efficacy of the design \citep{venable2016feds}. Given
that it is prohibitively expensive to evaluate the design a real-world
setting, we aim to minimize technological risks and maximize the
efficacy of the design by considering three examples of health data
platforms in LMICs, namely:

@@

\begin{itemize}
\tightlist
\item
  What is the level of openness of the implementation, specifically in
  terms of the three types of data sharing and platform-to-platform
  opennenss?
\item
  How are the core maintenance functions of the OHDP implemented, where
  we particularly focus on functions related to metadata management,
  data lineage and access control mechanisms?
\item
  Is the solution suitable for downward scalability, where we focus on
  the computer and storage requirements of the solution and assess the
  minimal requirements for on-site resources (computers, edge devices)
\end{itemize}

@@

\subsection{Open content}\label{open-content}

\begin{itemize}
\tightlist
\item
\end{itemize}

@@

\subsection{Jembi OpenHIM platform}\label{jembi-openhim-platform}

@@

\subsection{ONA OpenSRP 2}\label{ona-opensrp-2}

Continuing our evaluation of the extended OpenHIE architecture, we can
see a different flavor in the implementation driven by Ona. Ona is a
social enterprise that has pioneered the adoption of FHIR data standard
via the development of OpenSRP2, a FHIR-based data collection app built
using Google's FHIR SDK and focused on enabling offline-first workflows
for community-based care. OpenSRP 2 is a complete rewrite of the
original OpenSRP application, a global public good maintained by Ona and
deployed in XX countries worldwide.

OpenSRP2 applications are currently implemented in the field in three
countries (Uganda, Liberia, and Madagascar) in collaboration with local
Ministries of Health and with international donors such as UNICEF,
supporting a variety of different workflows including antenatal care
(ANC), postnatal Care (PNC), immunization, and last-mile logistics.
Besides the OpenSRP Android application and HAPI-FHIR backend, in each
of its projects Ona also implements a companion set of tools that
support analytics and various reporting needs.

\subsubsection{Requirements for data
sharing}\label{requirements-for-data-sharing}

Based on years of work in global health, Ona has learned that the data
stack implemented to support a national-scale implementation of its
FHIR-based application responds to the following requirements.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2857}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7143}}@{}}
\caption{Requirements and rationale for open health data platform
developed and used by ONA, based on OpenSRP 2
(\url{https://opensrp.io/}).}\label{tbl-ona-requirements}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Requirement
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Rationale
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Requirement
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Rationale
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Ingest data from multiple sources, both FHIR and non-FHIR based. & While
most health record data can be collected and aggregated in FHIR,
Ministries of Health rely on other data sources to govern their
operations. For example, operationalizing an immunization campaign
usually includes tracking against specific targets for locations to be
visited on specific days and number of children to immunize per day.
Such targets are often stored in spreadsheets or other applications
where the data is not FHIR. \\
Ingest data in batches. & Most data ingestion can happen in batches,
since Ona's applications are deployed in hard to reach areas where
connectivity is an issue. Data ingestion closer to real-time can be
relevant for disaster-response and other time-sensitive applications,
but this is not a priority. \\
Support national-scale data volumes. & A data store that can grow from
dozens to thousands of devices and where data can be aggregated up to
the national level, matching the scale of implementation of data
collection applications in the field. \\
Pre-compute complex business metrics. & Reporting on health systems
requires pre-computing complex metrics and often performing cohort
analyses to map trends in service provision. For example, understanding
quality of care for children requires computing metrics such as the
percentage of children fully immunized on schedule (i.e.~children 6-59
months that have received the set of vaccines required by the Ministry
of Health, and have received each of those vaccines within the expected
age-window). For Business Intelligence applications, calculating such a
vital metric cannot be performed at run time, to avoid long and
expensive queries. \\
Outbound integrations. & While aggregated data and reports should be
accessible by other applications such as BI platforms via pulls, there
should be an easy integration framework to push data to other
applications used by the Ministry of Health for other purposes, such as
DHIS2 for health systems management or RapidPro for communications with
program beneficiaries. \\
Open source and easily deployable in-country. & Given the extremely
sensitive nature of health data, it is paramount for governments to have
the flexibility to deploy the stack in various different environments,
both on premise and in private clouds. \\
\end{longtable}

The architecture Learning from experience in the field and internal
research and development, Ona has developed preferences for a specific
data stack responding to the aforementioned requirements.

{[}{[}graphic{]}{]}

Core toolings in the stack include: Data ingestion with Airbyte. Ona
uses Airbyte as the primary data ingestion tool, leveraging the wide
array of connectors that come standard with the application as well as a
dedicated suite of connectors developed internally by Ona, including
HAPI FHIR, RapidPro, Ona Data, Kobo Toolbox and others.\\
Data storage with Clickhouse. While different health projects have
varying requirements, Ona has found success in using Clickhouse as the
main analytics data store in its most recent implementations. Clickhouse
supports the scale required for analytics at a national level, as well
as the speed that enables cross-application integrations and more
real-time analytics. For example, in Madagascar Ona uses its reporting
suite to identify facilities with stock in need of maintenance and can
trigger the scheduling of a maintenance visit ad hoc. Data
transformation with dbt. Following global best practice, Ona leverages
dbt to segregate the data warehouse in different levels (staging, marts,
metrics), as well as pre-computing complex indicators for ease of
reporting and for transmissions into other systems. For example, in
Liberia Ona implements OpenSRP at community health worker level, but can
aggregate immunization data at facility level in the data warehouse and
then push quarterly summary metrics to DHIS2. No recommendation on
reporting / BI tooling. Ona recognizes that business users have their
own strong preferences for BI tooling, and some already have licenses
for specific software, so the architecture is flexible to provide easy
connections to different BI tools.

Evaluation

Evaluating the data stack, we see the following: Use of generic
best-of-breed tooling. Ona focused on utilizing Open HIE tools that are
widely adopted outside of the global health and development sectors.
This approach aims to provide assurance on two main fronts, the ability
to handle performance at scale and the long term dependability of the
tools, rather than relying on smaller projects with uncertain long term
funding or unproven implementations. Columnar data warehouse for
analytics. The scale of Ona's project requires the implementation of a
dedicated database for analytics. While original data can still be
stored as parquet or other file system, being able to ingest it into a
relational data store allows to create well defined indicators. Using
clickhouse as a tool helps and combine the need accuracy with the speed
of reporting as new data is ingested.\\
Strong emphasis on SQL. While Ona has tested and experimented with
FHIR-specific tooling, such as the definition of data projections using
sql-on-fhir, Ona found that relying on sql for coding business logic
remained the faster and most scalable approach.

In summary, for Ona building analytics with FHIR data looks similar to
building analytics with any other type of data. While FHIR provides a
clear and standard data model, managing information for most health
systems requires custom integration of data between different sources,
as well as computing indicators using business logic specific to the
needs of the local users. Building upon well established best-of-breed
tools allows Ona to implement FHIR applications at scale and provide
trusted analytics on top.

\subsection{PharmAccess demonstrator Momcare
programme}\label{pharmaccess-demonstrator-momcare-programme}

@@

Clearly, a more open and scaleable platform was required if MomCare was
to be implemented in more regions. This need led to a redesign of the
underlying technical infrastructure of the MomCare project. The
objectives of this work were in fact to demonstrate a solution design
that could support the first three types of data sharing. First, to
investigate the viability of using FHIR for bulk data sharing, MomCare
Tanzania was used a testbed to assess the complexity and effort required
to implement the facade pattern to integrate the legacy system into the
FHIR data standard. Using the longitudinal dataset from approximately 28
thousand patient records, FHIR transformations script were developed and
deployed using the mediator function of the IOL. The data was
transformed into 10 FHIR v4 resources and the conceptual data model of
the existing MomCare app could readily be transformed into the FHIR
standard using SQL and validated with a Python library
\citep{islam2023fhir}. The largest challenge during the transformation
process pertained to the absence of unique business identifiers for
patients and healthcare organizations. For patients, either the mobile
phone number or the healthcare insurance number was taken, depending on
availability. A combination of name, address and latitude/longitude
coordinates were used to uniquely identify organizations and locations,
as Tanzania does not have a system in place for this purpose.

@@

@@

@@

@@

We specifically address the notion of openness of OHDPs in LMICs in
terms of the design-related questions put forward by de Reuver at al.11:

\begin{itemize}
\tightlist
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different OHDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\item
  Answers/insights to above:

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ACID & Atomicity, Consistency, Isolation, and Durability \\
CLI & Command-line Interface \\
CR & Client Registry \\
ELK & Elasticsearch, Logstach and Kibana stack \\
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
OHDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
MPC & Multiparty Computation \\
PET & Privacy-enhancing technologies \\
SHR & Shared Health Record \\
\end{longtable}
